cardinalR: Generating interesting high-dimensional data structures

### 2025-10-05 paper-cardinalR.Rmd
Updated Table 12

An S-curve structure (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:scurve)"} else { "Figure \\@ref(fig:scurve-proj)"})`) simulates data that lies along a smooth, non-linear manifold.

For the S-curve structure, `gen_scurve(n, p)` (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:scurve) a"} else { "Figure \\@ref(fig:scurve-proj) a"})`), the $3\text{-}D$ geometry is constructed by introducing a latent parameter, $\theta \sim U\left(-3\pi/2, 3\pi/2\right)$. This parameter controls the curvature of the manifold. The first three dimensions form the S-curve structure: $$X_1 = \sin(\theta), \quad X_2 \sim U(0, 2), \quad X_3 = \text{sign}(\theta)(\cos(\theta) - 1)$$. This configuration creates a horizontally curled shape in $(X_1, X_3)$, with additional band thickness in the $X_2$ direction.

@@

Two specialized wrappers illustrate this idea. The function `gen_scurvehole()` generates an S-curve with a spherical hole by applying `gen_hole()` to the output of `gen_scurve()`. This structure has been used in prior diagnostic studies of NLDR methods [@wang2021], since it tests the ability of algorithms to capture non-linear manifolds that are not simply connected. The second wrapper, `gen_unifcubehole()`, generates uniformly sampled cube data with a hyperspherical hole. By embedding a hyperspherical void inside a convex high-dimensional structure, this creates non-convex regions that challenge algorithms in terms of separability and neighborhood preservation.

@@

```{r}
odd_shapes_tb <- tibble(fun = c("make_mobiusgau",
                                "make_multigau",
                                "make_curvygau",
                                "make_klink_circles",
                                "make_chain_circles",
                                "make_klink_curvycycle",
                                "make_chain_curvycycle",
                                "make_gaucircles",
                                "make_gaucurvycycle",
                                "make_onegrid",
                                "make_twogrid_overlap",
                                "make_twogrid_shift",
                                "make_shape_para",
                                "make_three_clust_"), 
                        exp = c(
    "Möbius-like cluster combined with a Gaussian.",
    "Multiple Gaussian clusters in high-dimensional space.",
    "Curvilinear cluster with a Gaussian cluster.",
    "K-link circular clusters (non-linear circular patterns).",
    "Chain-like circular clusters connected sequentially.",
    "K-link curvy cycle clusters (curvilinear loop structures).",
    "Chain-like curvy cycle clusters connected sequentially.",
    "Circular clusters with a Gaussian cluster in the middle.",
    "Curvy circular clusters with a Gaussian cluster in the middle.",
    "Single grid in two dimensions.",
    "Two overlapping grids.",
    "Two grids shifted relative to each other.",
    "Parallel shaped clusters.",
    "Three clusters with different shapes. (eg:- 01, 02, ..., 20)"))
```

>>>>

The S-curve structure (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:scurve)"} else { "Figure \\@ref(fig:scurve-proj)"})`) represents a smooth, non-linear manifold in 3D space. Using `gen_scurve(n, p)` (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:scurve) a"} else { "Figure \\@ref(fig:scurve-proj) a"})`), the geometry is defined by a latent parameter $\theta \sim U(-3\pi/2, 3\pi/2)$, which controls the curvature of the manifold. The first three coordinates form the S-shape:

$$
X_1 = \sin(\theta), \quad X_2 \sim U(0, 2), \quad X_3 = \text{sign}(\theta)(\cos(\theta) - 1),
$$

creating a horizontally curled pattern in $(X_1, X_3)$ with additional thickness along $X_2$.

@@

Two specialized wrappers illustrate this idea. The function `gen_scurvehole()` generates an S-curve with a spherical hole by applying `gen_hole()` to the output of `gen_scurve()`. This structure has been used in prior diagnostic studies of NLDR methods, since it tests the ability of algorithms to capture non-linear manifolds that are not simply connected. The second wrapper, `gen_unifcubehole()`, generates uniformly sampled cube data with a hyperspherical hole. By embedding a hyperspherical void inside a convex high-dimensional structure, this creates non-convex regions that challenge algorithms in terms of separability and neighborhood preservation.

@@

```{r}
odd_shapes_tb <- tibble(fun = c("make_mobiusgau",
                                "make_multigau",
                                "make_curvygau",
                                "make_klink_circles",
                                "make_chain_circles",
                                "make_klink_curvycycle",
                                "make_chain_curvycycle",
                                "make_gaucircles",
                                "make_gaucurvycycle",
                                "make_onegrid",
                                "make_twogrid_overlap",
                                "make_twogrid_shift",
                                "make_shape_para",
                                "make_three_clust_"), 
                        exp = c(
    "Möbius-like cluster combined with a Gaussian.",
    "Multiple Gaussian clusters in high-dimensional space.",
    "Curvilinear cluster with a Gaussian cluster.",
    "K-link circular clusters (non-linear circular patterns).",
    "Chain-like circular clusters connected sequentially.",
    "K-link curvy cycle clusters (curvilinear loop structures).",
    "Chain-like curvy cycle clusters connected sequentially.",
    "Circular clusters with a Gaussian cluster in the middle.",
    "Curvy circular clusters with a Gaussian in the middle.",
    "Single grid in two dimensions.",
    "Two overlapping grids.",
    "Two grids shifted relative to each other.",
    "Parallel shaped clusters.",
    "Three clusters with different shapes. (eg:- 01, 02, ..., 20)"))
```

### 2025-10-05 paper-cardinalR.tex
Updated the positions of the tables
Updated Table 12

An S-curve structure (Figure \ref{fig:scurve-proj}) simulates data that lies along a smooth, non-linear manifold.

For the S-curve structure, \texttt{gen\_scurve(n,\ p)} (Figure \ref{fig:scurve-proj} a), the \(3\text{-}D\) geometry is constructed by introducing a latent parameter, \(\theta \sim U\left(-3\pi/2, 3\pi/2\right)\). This parameter controls the curvature of the manifold. The first three dimensions form the S-curve structure: \[X_1 = \sin(\theta), \quad X_2 \sim U(0, 2), \quad X_3 = \text{sign}(\theta)(\cos(\theta) - 1)\]. This configuration creates a horizontally curled shape in \((X_1, X_3)\), with additional band thickness in the \(X_2\) direction.

@@

\subsubsection{Generate a Spherical or Hyperspherical Hole Within a Structure}\label{generate-a-spherical-or-hyperspherical-hole-within-a-structure}

@@

Two specialized wrappers illustrate this idea. The function \texttt{gen\_scurvehole()} generates an S-curve with a spherical hole by applying \texttt{gen\_hole()} to the output of \texttt{gen\_scurve()}. This structure has been used in prior diagnostic studies of NLDR methods \citep{wang2021}, since it tests the ability of algorithms to capture non-linear manifolds that are not simply connected. The second wrapper, \texttt{gen\_unifcubehole()}, generates uniformly sampled cube data with a hyperspherical hole. By embedding a hyperspherical void inside a convex high-dimensional structure, this creates non-convex regions that challenge algorithms in terms of separability and neighborhood preservation.

@@

\caption{\label{tab:odd-shape-tb-pdf}cardinalR multiple clusters generation functions}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{3.5cm}>{\raggedright\arraybackslash}p{8.5cm}}
\toprule
Function & Explanation\\
\midrule
make\_mobiusgau & Möbius-like cluster combined with a Gaussian.\\
make\_multigau & Multiple Gaussian clusters in high-dimensional space.\\
make\_curvygau & Curvilinear cluster with a Gaussian cluster.\\
make\_klink\_circles & K-link circular clusters (non-linear circular patterns).\\
make\_chain\_circles & Chain-like circular clusters connected sequentially.\\
make\_klink\_curvycycle & K-link curvy cycle clusters (curvilinear loop structures).\\
make\_chain\_curvycycle & Chain-like curvy cycle clusters connected sequentially.\\
make\_gaucircles & Circular clusters with a Gaussian cluster in the middle.\\
make\_gaucurvycycle & Curvy circular clusters with a Gaussian cluster in the middle.\\
make\_onegrid & Single grid in two dimensions.\\
make\_twogrid\_overlap & Two overlapping grids.\\
make\_twogrid\_shift & Two grids shifted relative to each other.\\
make\_shape\_para & Parallel shaped clusters.\\
make\_three\_clust\_ & Three clusters with different shapes. (eg:- 01, 02, ..., 20)\\
\bottomrule
\end{tabular}
\end{table}

>>>>

The S-curve structure (Figure \ref{fig:scurve-proj}) represents a smooth, non-linear manifold in 3D space. Using \texttt{gen\_scurve(n,\ p)} (Figure \ref{fig:scurve-proj} a), the geometry is defined by a latent parameter \(\theta \sim U(-3\pi/2, 3\pi/2)\), which controls the curvature of the manifold. The first three coordinates form the S-shape:

\[
X_1 = \sin(\theta), \quad X_2 \sim U(0, 2), \quad X_3 = \text{sign}(\theta)(\cos(\theta) - 1),
\]

creating a horizontally curled pattern in \((X_1, X_3)\) with additional thickness along \(X_2\).

@@

\subsubsection{Generate a spherical or hyperspherical hole within a structure}\label{generate-a-spherical-or-hyperspherical-hole-within-a-structure}

@@

Two specialized wrappers illustrate this idea. The function \texttt{gen\_scurvehole()} generates an S-curve with a spherical hole by applying \texttt{gen\_hole()} to the output of \texttt{gen\_scurve()}. This structure has been used in prior diagnostic studies of NLDR methods, since it tests the ability of algorithms to capture non-linear manifolds that are not simply connected. The second wrapper, \texttt{gen\_unifcubehole()}, generates uniformly sampled cube data with a hyperspherical hole. By embedding a hyperspherical void inside a convex high-dimensional structure, this creates non-convex regions that challenge algorithms in terms of separability and neighborhood preservation.

@@

\caption{\label{tab:odd-shape-tb-pdf}cardinalR multiple clusters generation functions}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{3.5cm}>{\raggedright\arraybackslash}p{8.5cm}}
\toprule
Function & Explanation\\
\midrule
make\_mobiusgau & Möbius-like cluster combined with a Gaussian.\\
make\_multigau & Multiple Gaussian clusters in high-dimensional space.\\
make\_curvygau & Curvilinear cluster with a Gaussian cluster.\\
make\_klink\_circles & K-link circular clusters (non-linear circular patterns).\\
make\_chain\_circles & Chain-like circular clusters connected sequentially.\\
make\_klink\_curvycycle & K-link curvy cycle clusters (curvilinear loop structures).\\
make\_chain\_curvycycle & Chain-like curvy cycle clusters connected sequentially.\\
make\_gaucircles & Circular clusters with a Gaussian cluster in the middle.\\
make\_gaucurvycycle & Curvy circular clusters with a Gaussian in the middle.\\
make\_onegrid & Single grid in two dimensions.\\
make\_twogrid\_overlap & Two overlapping grids.\\
make\_twogrid\_shift & Two grids shifted relative to each other.\\
make\_shape\_para & Parallel shaped clusters.\\
make\_three\_clust\_ & Three clusters with different shapes. (eg:- 01, 02, ..., 20)\\
\bottomrule
\end{tabular}
\end{table}
